Modal Abstractions for Virtualizing Memory Addresses

### 2024-07-09 programlogic.tex
fix a few typos, note critical missing content in figure 2?

% The idea that the truth of an assertion is relative to an address space has far-reaching consequences.
To support making assertions depend on a choice of address space, we work entirely in a pointwise lifting of Iris's base BI logic,
essentially working with separation logic assertions indexed by a choice of page table root as a $\mathcal{W}_{64}$, which we call $\textsf{vProp }\Sigma$:\footnote{
  \iris experts may notice our \lstinline|-b>| resembles another variety of pointwise lifting that already exists in the main \iris
  distribution~\cite{dang2019rustbelt,dang2022compass}. This is essentially true, but the standard formalization does not work for index types which
  are indexed in the sense of \coq's metatheory, as our \lstinline|word n| types are.
}
\begin{lstlisting}[language=Coq]
  Definition vProp  $\Sigma$ : bi := word 64 -b> iPropI  $\Sigma$.
\end{lstlisting}
This is the (\textsc{Coq}) type of assertions in our logic.
Most constructs in Iris's base logic are defined with respect to any BI-algebra (of \textsc{Coq} type \lstinline|bi|), so automatically
carry over to our derived logic.
However, be must still build up from existing Iris primitives to provide new primitives that depend on the address space --- primarily the notion
of virtual points-to.

@@

\subsection{An Overly-Restrictive Definition for Virtual Memory Addressing}
\label{sec:overly-restrictive}
A natural definition for a virtual points-to that depends on the current address space is to
require that in order for a virtual address \textsf{va} to point to a value \textsf{v}, the assertion contains
partial ownership of the physical memory involved in the page table walk that would translate \textsf{va} to
its backing physical location --- with locations existentially quantified since a virtual points-to should not assert
\emph{which} locations are accessed in a page table walk.
Figure \ref{fig:strongvirtualpointsto} gives such a definition.
It asserts the existence of for page table entries, one at each translation level, and via \lstinline|L4_L1_PointsTo|
asserts from the current \lstinline|cr3| value, taking the offsets from \textsf{va} (as in Figure \ref{fig:pagetables})
to index each level, finding the existentially quantified entries in each level, 
ending with the L1 entry, whose physical page base is added to the page offset of \textsf{va} to obtain the physical address
 --- and there (\lstinline|addr_L1(va,l1e)|) in physical memory, is the value.
Most of the definition lives directly in \textsf{vProp}, using the separation logic structure lifted from Iris's \textsf{iProp}.

@@

  We take separate the physical page-table-walk  
  from the virtual pointsto relation, replacing it with ghost state that merely guarantees the address translation would succeed.
  Iris includes a construction which permits tracking a ghost map 
  from virtual addresses to the physical addresses they translate to, and has for each key in the map (i.e.,
  each virtual address) a token. Updating the virtual-to-physical (ghost) mapping for a virtual address $va$
  requires full ownership of the token. Then the \emph{virtual memory manager's invariant} ensures that for each $va\mapsto pa$ mapping
  in this map, there are \emph{physical} resources sufficient to ensure that address translation for $va$
  will resolve in the hardware to $pa$ --- $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$.
  This allows modification of the page tables themselves as long as the changes do not invalidate overall virtual-to-physical
  translation.
  Consistency follows from replacing the page table walk with ownership of the ghost map token for the relevant ghost map
  entry, preventing the overall translation from being changed separately from the virtual points-to.

>>>>

% The idea that the truth of an assertion is relative to an address space has far-reaching consequences.
To support making assertions depend on a choice of address space, we work entirely in a pointwise lifting of Iris's base BI logic,
essentially working with separation logic assertions indexed by a choice of page table root as a $\mathcal{W}_{64}$, which we call $\textsf{vProp }\Sigma$:\footnote{
  \iris experts may notice our \lstinline|-b>| resembles another variety of pointwise lifting that already exists in the main \iris
  distribution~\cite{dang2019rustbelt,dang2022compass}. This is essentially true, but the standard formalization does not work for index types which
  are indexed in the sense of \coq's metatheory, as our \lstinline|word n| types are.
}
\begin{lstlisting}[language=Coq]
  Definition vProp  $\Sigma$ : bi := word 64 -b> iPropI  $\Sigma$.
\end{lstlisting}
This is the (\textsc{Coq}) type of assertions in our logic.
Most constructs in Iris's base logic are defined with respect to any BI-algebra (of \textsc{Coq} type \lstinline|bi|), so automatically
carry over to our derived logic.
However, we must still build up from existing Iris primitives to provide new primitives that depend on the address space --- primarily the notion
of virtual points-to.

@@

\subsection{An Overly-Restrictive Definition for Virtual Memory Addressing}
\label{sec:overly-restrictive}
A natural definition for a virtual points-to that depends on the current address space is to
require that in order for a virtual address \textsf{va} to point to a value \textsf{v}, the assertion contains
partial ownership of the physical memory involved in the page table walk that would translate \textsf{va} to
its backing physical location --- with locations existentially quantified since a virtual points-to should not assert
\emph{which} locations are accessed in a page table walk.
Figure \ref{fig:strongvirtualpointsto} gives such a definition.
It asserts the existence of four page table entries, one at each translation level, and via \lstinline|L4_L1_PointsTo|
asserts from the current \lstinline|cr3| value, taking the offsets from \textsf{va} (as in Figure \ref{fig:pagetables})
to index each level, finding the existentially quantified entries in each level, 
ending with the L1 entry, whose physical page base is added to the page offset of \textsf{va} to obtain the physical address
 --- and there (\lstinline|addr_L1(va,l1e)|) in physical memory, is the value.
Most of the definition lives directly in \textsf{vProp}, using the separation logic structure lifted from Iris's \textsf{iProp}.
\todo[inline]{Colin notes: the \lstinline|addr_L1| mentioned in the paragraph above doesn't appear in Figure 2. paddr isn't mentioned
in the definition of \lstinline|L4_L1_PointsTo| in the figure. Shouldn't there be a pure \lstinline|addr_L1(va,l1e)=paddr| in there?
}

@@

  We separate the physical page-table-walk  
  from the virtual pointsto relation, replacing it with ghost state that merely guarantees the address translation would succeed.
  Iris includes a construction which permits tracking a ghost map 
  from virtual addresses to the physical addresses they translate to, and has for each key in the map (i.e.,
  each virtual address) a token. Updating the virtual-to-physical (ghost) mapping for a virtual address $va$
  requires full ownership of the token. Then the \emph{virtual memory manager's invariant} ensures that for each $va\mapsto pa$ mapping
  in this map, there are \emph{physical} resources sufficient to ensure that address translation for $va$
  will resolve in the hardware to $pa$ --- $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$.
  This allows modification of the page tables themselves as long as the changes do not invalidate overall virtual-to-physical
  translation.
  Consistency follows from replacing the page table walk with ownership of the ghost map token for the relevant ghost map
  entry, preventing the overall translation from being changed separately from the virtual points-to.
